---
title: "Life Expectancy - Transformations and Weighted Least Squares"
author: "Kasia Krueger"
date: "10/08/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('car')
library('olsrr')
#library('rgl')
setwd("~/Personal/School/MA4710/Project")
lifedata <- read.csv("Life Expectancy Data_2015.csv",row.names=1)

polio.vacc <- lifedata$Polio
lifedata$polio.vacc <- cut(polio.vacc,3,labels = c("1", "2", "3"))

HIV.AIDS <- as.numeric(lifedata$HIV.AIDS)  
lifedata$HIV.AIDS<- cut(HIV.AIDS,3,labels = c("1", "2", "3"))

```



# Week 5

## Cleaning up data
Removing high influential points (errors in years of schooling as found in Week 3 Regression diagnostics.)
```{r}
lifedata <- lifedata[-which(rownames(lifedata) == "Burkina Faso"),] 
#Possible error in years of schooling (26)
lifedata <- lifedata[-which(rownames(lifedata) == "Equatorial Guinea"),] 
#Possible error in years of schooling (32)
lifedata <- lifedata[-which(rownames(lifedata) == "Cuba"),] 
#Error in years of schooling (92)
lifedata <- lifedata[-which(rownames(lifedata) == "Eritrea"),] 
# Error in years of schooling (255)
lifedata <- lifedata[-which(rownames(lifedata) == "India"),] 
#High cook's distance + possible error in infant deaths (800/1000)
```

```{r}
lifedata$polio.vacc <- as.factor(lifedata$polio.vacc)
lifedata$HIV.AIDS <- as.factor(lifedata$HIV.AIDS)
lifedata$Status <- as.factor(lifedata$Status)

levels(lifedata$polio.vacc) <- c('low', 'medium', 'high')
levels(lifedata$HIV.AIDS) <- c('low', 'medium', 'high')
```

## Base Model 
```{r}
life.lm <- lm(Life.expectancy ~ Adult.Mortality 
              + Schooling 
              + BMI + GDP 
              + infant.deaths 
              + polio.vacc 
              + HIV.AIDS 
              + Status, lifedata)
scatterplotMatrix(~Adult.Mortality+BMI+GDP+Schooling+infant.deaths,lifedata, smooth=FALSE)

summary(life.lm)
hist(rstandard(life.lm), probability=TRUE)
par(mfrow=c(2,2))
plot(life.lm)
shapiro.test(rstandard(life.lm))
ncvTest(life.lm)
residualPlots(life.lm, type="rstandard")
```

# Summary: Base Model and Diagnostics

The scatterplot matrix shows there are non-linear relationship issues with infant deaths, as well as with GDP. Adult mortality and years of schooling appear to have a strong linear relationship, as well as with BMI and years of schooling. 

## Homoscedastcity

The standardized residuals plots and scale-location plot show that the homoscedasticity assumptions appear to be met, as the residuals are scattered evenly around 0. The Breusch-Pagan test has a p-value of 0.15, suggesting this test matches the visual of homoscedasticity. 

## Normality
The histogram appears to be left-skewed with large outliers up to -4. The QQ-plot appears to have heavy tails and that the data set is not normal. The small p-value in the Shapiro-Wilk test confirms this. However, the observations in GDP appear to be right-skewed. 

## Linearity 
The linearity assumption appears to be met for the coefficients, with the exception of Adult.mortality, which appears to have a quadratic response. The residuals vs. fitted values plot appears to be nearly horizontal, indicating linearity in the model.  

## Influential Observations

There do not appear to be influential observations. The 'Residuals vs Leverage' plot shows no observations over 0.5. 


To address the linearity assumptions in the model, we test the response and the predictor variables

# Addressing Assumption Violations
## Test response transformation
```{r}
invResPlot(life.lm)
```
### Lambda near 1 suggests no transformation is needed on the response.


## Test adult mortality predictor transformation
```{r}
invTranPlot(Life.expectancy~Adult.Mortality, lifedata)$lambda
```
### Lambda near 2 suggests adding a quadratic term, I(X^2), to the model
```{r}
life.lm2 <- lm(Life.expectancy ~ Adult.Mortality+ I(Adult.Mortality^2) + Schooling + BMI + GDP + infant.deaths + polio.vacc + HIV.AIDS, lifedata)
summary(life.lm2)
residualPlots(life.lm2, type="rstandard")
```
Adding a quadratic term to the model appears to improve the linearity assumption for adult mortality. 

## Test schooling predictor transformation
```{r}
invTranPlot(Life.expectancy~Schooling, lifedata)$lambda
```
### Lambda near 1 suggests no transformation is needed on the predictor

## Test BMI predictor transformation
```{r}
invTranPlot(Life.expectancy~BMI, lifedata)$lambda
```

### Lambda near 2 suggests adding a quadratic term, I(X^2), to the model
```{r}
life.lm3 <- lm(Life.expectancy ~ Adult.Mortality+I(Adult.Mortality^2) + Schooling + BMI+I(BMI^2) + GDP + infant.deaths + polio.vacc + HIV.AIDS, lifedata)
summary(life.lm3)
residualPlots(life.lm3, type="rstandard")

```
Adding a quadratic term to the model appears to improve the linearity assumption for adult BMI.


## Test GDP predictor transformation
```{r}
invTranPlot(Life.expectancy~GDP, lifedata)$lambda
```
### Lambda near 0 suggests trying a logarithmic transformation on the predictor
```{r}
life.lm4 <- lm(Life.expectancy ~ Adult.Mortality+I(Adult.Mortality^2) + Schooling + BMI+I(BMI^2) + log(GDP) + infant.deaths + polio.vacc + HIV.AIDS, lifedata)
summary(life.lm4)
residualPlots(life.lm4, type="rstandard")

```

Test not done on infant deaths since there are values of zero. 

## Removing insignificant predictors from model and adding interactions to the model:
```{r}
life.reduced <- update(life.lm4, ~.   
                       + I(Adult.Mortality^2):polio.vacc
                       + I(Adult.Mortality^2):Status
                       + I(Adult.Mortality^2):HIV.AIDS
                       + Schooling:polio.vacc
                       + Schooling:Status
                       + Schooling:HIV.AIDS
                       + HIV.AIDS:polio.vacc
                       - Adult.Mortality
                       - BMI
                       - I(BMI^2) 
                       - log(GDP) 
                       - infant.deaths 
                       - polio.vacc
                       - HIV.AIDS)

summary(life.reduced)

```
###  Small p-value (< 2.2e-16) suggests interaction effects need to stay in the model. 
### Now, we choose which interaction effects need to stay in the model.

```{r}
life.reduced1 <- update(life.reduced, ~ . - I(Adult.Mortality^2):polio.vacc)
anova(life.reduced1, life.reduced)$`Pr(>F)`
```

###  Large p-value (0.305786) suggests we can remove I(Adult.Mortality^2):polio.vacc from model.

```{r}
life.reduced2 <- update(life.reduced1, ~ . - I(Adult.Mortality^2):HIV.AIDS)
anova(life.reduced2, life.reduced1)$`Pr(>F)`
```

###  Small p-value (7.937699e-05) suggests we need to keep I(Adult.Mortality^2):HIV.AIDS in the model.

```{r}
life.reduced3 <- update(life.reduced1, ~ . - Schooling:Status)
anova(life.reduced3, life.reduced1)$`Pr(>F)`
```

###  Small p-value (0.0005545418) suggests we need to keep Schooling:Status in the model.

```{r}
life.reduced4 <- update(life.reduced1, ~ . - Schooling:HIV.AIDS)
anova(life.reduced4, life.reduced1)$`Pr(>F)`
```

###  Small p-value (6.212583e-06) suggests we need to keep Schooling:HIV.AIDS from model.

```{r}
life.reduced5 <- update(life.reduced1, ~ . - HIV.AIDS:polio.vacc)
anova(life.reduced5, life.reduced1)$`Pr(>F)`
```

###  Large p-value (0.312771) suggests we can remove HIV.AIDS:polio.vacc from model.

```{r}
life.reduced6 <- update(life.reduced5, ~ . - Schooling:polio.vacc)
anova(life.reduced6, life.reduced5)$`Pr(>F)`

```

###  Small p-value (0.006823202) suggests we need to keep Schooling:polio.vacc in the model.

# Diagnostics for Final Model:
```{r}
summary(life.reduced5)
par(mfrow=c(2,2))
plot(life.reduced5)
shapiro.test(rstandard(life.reduced5))
ncvTest(life.reduced5)
par(mfrow=c(1,1))
hist(rstandard(life.reduced5), probability=TRUE)
residualPlots(life.reduced5, id=TRUE, quadratic=TRUE, type='rstandard', tests=FALSE)
```


# Summary: Final Model and Diagnostics

## Homoscedastcity

The standardized residuals plots and scale-location plot show that the homoscedasticity assumptions appear to be met, as the residuals are scattered evenly around 0. The Breusch-Pagan test has a p-value of 0.83359, suggesting this test matches the visual of homoscedasticity. 

## Normality
The histogram looks normally distributed. The QQ-plot appears to be normal with no heavy tails. The large p-value in the Shapiro-Wilk test (p-value = 0.4801) confirms this.

## Linearity 
The linearity assumption appears to be met for the coefficients. The residuals vs. fitted values plot appears to be nearly horizontal, indicating linearity in the model.  

## Influential Observations

There do not appear to be influential observations. The 'Residuals vs Leverage' plot shows no observations over 0.5. 

## R^2^,  adjusted R^2^, sigma-hat
## Base Model
```{r}
summary(life.lm)$r.squared  
summary(life.lm)$adj.r.squared 
summary(life.lm)$sigma
```

## Final Model
```{r}
summary(life.reduced5)$r.squared
summary(life.reduced5)$adj.r.squared
summary(life.reduced5)$sigma
```
 
The model improved overall from the base model, with an improvement to the   adjusted R^2^ value, meaning there was not a "penalty" for adding the additional interactions to the model, as well as a reduction of the standard error, resulting in a more accurate model. R^2^ improved as well, solely due to the addition of the interactions. 
