---
title: "Life Expectancy - Collinearity and Variable Selection"
author: "Kasia Krueger"
date: "10/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('car')
library('olsrr')
#library('rgl')
setwd("~/Personal/School/MA4710/Project")
lifedata <- read.csv("Life Expectancy Data_2015.csv",row.names=1)

polio.vacc <- lifedata$Polio
lifedata$polio.vacc <- cut(polio.vacc,3,labels = c("1", "2", "3"))

HIV.AIDS <- as.numeric(lifedata$HIV.AIDS)  
lifedata$HIV.AIDS<- cut(HIV.AIDS,3,labels = c("1", "2", "3"))

```



# Week 7

## Cleaning up data
Removing high influential points (errors in years of schooling as found in Week 3 Regression diagnostics.)
```{r}
lifedata <- lifedata[-which(rownames(lifedata) == "Burkina Faso"),] 
#Possible error in years of schooling (26)
lifedata <- lifedata[-which(rownames(lifedata) == "Equatorial Guinea"),] 
#Possible error in years of schooling (32)
lifedata <- lifedata[-which(rownames(lifedata) == "Cuba"),] 
#Error in years of schooling (92)
lifedata <- lifedata[-which(rownames(lifedata) == "Eritrea"),] 
# Error in years of schooling (255)
lifedata <- lifedata[-which(rownames(lifedata) == "India"),] 
#High cook's distance + possible error in infant deaths (800/1000)
```

```{r}
lifedata$polio.vacc <- as.factor(lifedata$polio.vacc)
lifedata$HIV.AIDS <- as.factor(lifedata$HIV.AIDS)
lifedata$Status <- as.factor(lifedata$Status)

levels(lifedata$polio.vacc) <- c('low', 'medium', 'high')
levels(lifedata$HIV.AIDS) <- c('low', 'medium', 'high')
```

# Full Model 
```{r}
life.lm <- lm(Life.expectancy ~ Adult.Mortality 
              + Schooling 
              + BMI + GDP 
              + infant.deaths 
              + polio.vacc 
              + HIV.AIDS 
              + Status, lifedata)


summary(life.lm)
summary(life.lm)$sigma
```


## Calculate VIFS
```{r}
car::vif(life.lm)
```


## Condition number
```{r}
life.pc <- prcomp(~ Adult.Mortality 
              + Schooling 
              + BMI 
              + GDP 
              + infant.deaths 
              - Life.expectancy, lifedata, scale=TRUE)
PC.sdev <- life.pc$sdev
PC.sdev[1] / PC.sdev
```

# Summary: Collinearity in Full Model
The full model did not detect any issues with collinearity. Next, we explore this topic more during the variable selection process. 

# Best p-Variable Models
```{r}
life.rs <- leaps::regsubsets(Life.expectancy ~ Adult.Mortality 
              + Schooling 
              + BMI + GDP 
              + infant.deaths 
              + polio.vacc 
              + HIV.AIDS 
              + Status, lifedata)
              
summary(life.rs)

```
* Best 1-variable model is Y = $\beta_0$+$\beta_2$$X_2$
* Best 2-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$
* Best 3-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$
* Best 4-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_{10}$$X_{10}$
* Best 5-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_{10}$$X_{10}$
* Best 6-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_{10}$$X_{10}$
* Best 7-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_9$$X_9$ + $\beta_{10}$$X_{10}$
* Best 8-variable model is Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_3$$X_3$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_9$$X_9$ + $\beta_{10}$$X_{10}$

## Patterns:  

* $X_2$ *Schooling* is in all the models.
* $X_1$ *Adult Mortality* is in all the models containing two or more variables.
* $X_7$ *Polio Vaccine - High* is in all the models containing three or more variables.
* $X_{10}$ *Status - Developing* is in all the models containing four or more variables.

# Forward selection, backward elimination, and stepwise selection

## Forward AIC
```{r}
mod.null <- lm(Life.expectancy ~ 1, lifedata) # null model with no predictors
step(mod.null, scope=formula(life.lm), direction='forward', trace=FALSE)
```

* The best model using Forward AIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ +  $\beta_9$$X_9$ + $\beta_{10}$$X_{10}$

## Backward AIC
```{r}
step(life.lm, direction='backward', trace=FALSE)
```

* The best model using Backward AIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ +  $\beta_9$$X_9$ + $\beta_{10}$$X_{10}$

## Stepwise AIC
```{r}
step(life.lm, direction='both', trace=FALSE)
```

* The best model using Stepwise AIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ +  $\beta_9$$X_9$ + $\beta_{10}$$X_{10}$

All three methods produced the same model but it seems overfit. This process will be repeated using stepwise elimination with BIC, which has a strong penalty associated with addin too many terms. 

## Forward BIC
```{r}
mod.null <- lm(Life.expectancy ~ 1, lifedata) # null model with no predictors
step(mod.null, scope=formula(life.lm), direction='forward', trace=FALSE, k=log(nrow(lifedata)))
```

* The best model using Forward BIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ +  $\beta_7$$X_7$ + $\beta_8$$X_8$ 

## Backward BIC
```{r}
step(life.lm, direction='backward', trace=FALSE, k=log(nrow(lifedata)))
```
* The best model using Backward BIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ +  $\beta_7$$X_7$ + $\beta_8$$X_8$ 

## Stepwise BIC
```{r}
step(life.lm, direction='both', trace=FALSE, k=log(nrow(lifedata)))
```
* The best model using Stepwise BIC is 
Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ +  $\beta_7$$X_7$ + $\beta_8$$X_8$ 

Again, all three methods produced the same model but with fewer predictors/variables using the BIC criteria.  

# Best subsets model selection
```{r}
par(mfrow=c(1,2))
plot(life.rs)
plot(life.rs, scale='adjr2')
plot(life.rs, scale='Cp')
```

* The model with lowest BIC: Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$  
* The model with highest R^2^_adj_: Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_4$$X_4$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_{10}$$X_{10}$  
* The model with lowest C{p}: Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_{10}$$X_{10}$

```{r}
subsets(life.rs, statistic='adjr2',  ylim=c(.8,.86), min.size=2, legend=FALSE)

subsets(life.rs, statistic='cp', xlim=c(3,8), ylim=c(4.5,8), min.size=2, legend=FALSE)
abline(a=1, b=1, col='red', lty='dashed', lwd=2)
```
## Summary: Subsets

The R^2^_adj_ plot is difficult to interpret, since many of the models have R^2^_adj_ from 0.83-0.85, as seen in the  _plot_ for R^2^_adj_.

From the cp plot, the first model which crosses below the reference line is also the model with the lowest C{p}: Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_8$$X_8$ + $\beta_{10}$$X_{10}$

# Variable selection procedures

## Lowest Cp Model
Using the model with the lowest C{p}:
```{r}
life_cp <- lm(Life.expectancy ~ 
            Adult.Mortality
              + Schooling 
              + polio.vacc
              + HIV.AIDS 
              + Status, lifedata)
              
summary(life_cp) 
summary(life_cp)$sigma

```



```{r}
vif(life_cp)

```


```{r}
par(mfrow=c(2,2))
plot(life_cp)

```


```{r}
res.cp <- rstandard(life_cp)
shapiro.test(res.cp)
car::ncvTest(life_cp)
par(mfrow=c(1,1))
hist(res.cp)
residualPlots(life_cp, type='rstandard', quadratic=FALSE)

```
## Summary: Lowest Cp Model 

The summary statistics look good, with most predictors being influential and a R^2^_adj_ of 0.8435 and sigma of 3.27.There are no issues with collinearity as all VIFs are less than 10. 

The plots do not indiate any issues with normality, lineairtiy or homoscedasticity. There is one influential point (Zambia). 

The Breusch-Pagan test indicates there is an issue with heteroscadisity that was not seen in the plots. The histogram of residuals is unfortunately left-skewed, with some high outliers. 

## Lowest BIC Model

Next we will try using the model (tied) with lowest BIC: Y = $\beta_0$ + $\beta_1$$X_1$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_{10}$$X_{10}$ 
```{r}
 
life_BIC <- lm(Life.expectancy ~ 
               Adult.Mortality
              + Schooling 
              + polio.vacc
              + Status, lifedata)
              
summary(life_BIC) 
summary(life_BIC)$sigma 
par(mfrow=c(2,2))
plot(life_BIC)
res.BIC <- rstandard(life_BIC)
shapiro.test(res.BIC)
car::ncvTest(life_BIC)
par(mfrow=c(1,1))
hist(res.BIC)
residualPlots(life_BIC, type='rstandard', quadratic=FALSE)

```
## Summary: Lowest BIC Model 

The summary statistics look good, with most predictors being influential and a R^2^_adj_ of 0.8416 and sigma of 3.99. 

The Breusch-Pagan test improved significantly, as did the influntial point (Zambia), which has moved back under a Cook's distance of 0.5. 

The histogram of residuals has improved with fewer outliers, but still appears slightly left-skewed.  

## Lowest BIC Model with Quadratic Transformation

Next we will try using the same model with the lowest BIC, but with a quadratic predictor: Y = $\beta_0$ + $\beta_1$$X_1^2$ + $\beta_2$$X_2$ + $\beta_7$$X_7$ + $\beta_{10}$$X_{10}$ 
```{r}
 
life_BIC2 <- lm(Life.expectancy ~ 
               Adult.Mortality^2
              + Schooling 
              + polio.vacc
              + Status, lifedata)
              
summary(life_BIC2)
summary(life_BIC2)$sigma
par(mfrow=c(2,2))
plot(life_BIC2)
res.BIC2 <- rstandard(life_BIC2)
shapiro.test(res.BIC2)
car::ncvTest(life_BIC2)
par(mfrow=c(1,1))
hist(res.BIC2)
residualPlots(life_BIC2, type='rstandard', quadratic=FALSE)
```

## Summary: Lowest BIC Model with Quadratic Transformation 

The residuals have improved a bit but there are still some significant outliers in the histogram of residuals (>|3|). Sigma has improved a bit to 3.35. Adding in some interaction terms like in Week 4 could help improve the model even further.

# Final model

Using the lowest BIC model coefficients, along with transformations and interactions found previously to further explore the "best" model for this data. 

```{r}
life.final <- lm(Life.expectancy ~ 
I(Adult.Mortality^2) 
+ Schooling 
+ I(Adult.Mortality^2):Status 
+ I(Adult.Mortality^2):HIV.AIDS 
+ Schooling:polio.vacc 
+ Schooling:Status 
+ Schooling:HIV.AIDS,
data = lifedata)
               
summary(life.final)
summary(life.final)$sigma
par(mfrow=c(2,2))
plot(life.final)
res.final <- rstandard(life.final)
shapiro.test(res.final)
car::ncvTest(life.final)
par(mfrow=c(1,1))
hist(res.final)
residualPlots(life.final, type='rstandard', quadratic=FALSE)
influencePlot(life.final)

```
# Summary: Final Model 

Y = $\beta_0$ 
+ $\beta_1$$X_1^2$ 
+ $\beta_2$$X_2$ 
+ $\beta_1$$X_1^2$:$\beta_{10}$$X_{10}$
+ $\beta_1$$X_1^2$:$\beta_{8}$$X_{8}$
+ $\beta_1$$X_1^2$:$\beta_{9}$$X_{9}$
+ $\beta_2$$X_2$*$\beta_{6}$$X_{7}$
+ $\beta_2$$X_2$*$\beta_{7}$$X_{7}$
+ $\beta_2$$X_2$*$\beta_{10}$$X_{10}$
+ $\beta_2$$X_2$*$\beta_{8}$$X_{8}$
+ $\beta_2$$X_2$*$\beta_{9}$$X_{9}$

The final model using the lowest BIC as well as the transformations and interactions from Week 5 produced the best results yet. It produced the highest R^2^_adj_ value yet: 0.8882, and lowest sigma yet: 2.78. 

Most interestingly is that the model includes the variables most commonly found in the best p-variable models observed during the variable selection procedures. The variables that occurred most often in the models ($X_2$ *Schooling*, $X_1$ *Adult Mortality*, $X_7$ *Polio Vaccine - High*, and $X_{10}$ *Status - Developing*) are all predictors used in the final model.

The assumptions for the residuals in the final model are also the best found in the 7 weeks of the course. The Breusch-Pagan test indicates homoscedasticity is met, and the normality assumption is met using the Shapiro-Wilk test. The linearity assumption is met; the only influential point is "Zambia" but it does not appear to be an outlier