---
title: "Life Expectancy - Collinearity and Variable Selection"
author: "Kasia Krueger"
date: "10/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('car')
library('olsrr')
#library('rgl')
setwd("~/Personal/School/MA4710/Project")
lifedata <- read.csv("Life Expectancy Data_2015.csv",row.names=1)

polio.vacc <- lifedata$Polio
lifedata$polio.vacc <- cut(polio.vacc,3,labels = c("1", "2", "3"))

HIV.AIDS <- as.numeric(lifedata$HIV.AIDS)  
lifedata$HIV.AIDS<- cut(HIV.AIDS,3,labels = c("1", "2", "3"))

```



# Week 7

## Cleaning up data
Removing high influential points (errors in years of schooling as found in Week 3 Regression diagnostics.)
```{r}
lifedata <- lifedata[-which(rownames(lifedata) == "Burkina Faso"),] 
#Possible error in years of schooling (26)
lifedata <- lifedata[-which(rownames(lifedata) == "Equatorial Guinea"),] 
#Possible error in years of schooling (32)
lifedata <- lifedata[-which(rownames(lifedata) == "Cuba"),] 
#Error in years of schooling (92)
lifedata <- lifedata[-which(rownames(lifedata) == "Eritrea"),] 
# Error in years of schooling (255)
lifedata <- lifedata[-which(rownames(lifedata) == "India"),] 
#High cook's distance + possible error in infant deaths (800/1000)
```

```{r}
lifedata$polio.vacc <- as.factor(lifedata$polio.vacc)
lifedata$HIV.AIDS <- as.factor(lifedata$HIV.AIDS)
lifedata$Status <- as.factor(lifedata$Status)

levels(lifedata$polio.vacc) <- c('low', 'medium', 'high')
levels(lifedata$HIV.AIDS) <- c('low', 'medium', 'high')
```

## Base Model 
```{r}
life.lm <- lm(Life.expectancy ~ Adult.Mortality 
              + Schooling 
              + BMI + GDP 
              + infant.deaths 
              + polio.vacc 
              + HIV.AIDS 
              + Status, lifedata)


summary(life.lm)

```


```
# Calculate VIFS
```{r}
car::vif(life.lm)
```


# Condition number
```{r}
life.pc <- prcomp(~ Adult.Mortality 
              + Schooling 
              + BMI 
              + GDP 
              + infant.deaths 
              - Life.expectancy, lifedata, scale=TRUE)
PC.sdev <- life.pc$sdev
PC.sdev[1] / PC.sdev
```

The base model did not detect any issues with collinearity. Next, we explore this topic more during the variable selection process. 

#Stepwise
```{r}

step(life.lm, direction='both')
step(life.lm, direction='both', k=log(nrow(lifedata)))

life.mod <- lm(formula = Life.expectancy ~ Adult.Mortality + Schooling + polio.vacc, data = lifedata)
summary(life.mod)
vif(life.mod)
residualPlots(life.mod, quadratic=FALSE, id=TRUE)

life.rs <- regsubsets(Life.expectancy ~ Adult.Mortality 
              + Schooling 
              + BMI + GDP 
              + infant.deaths 
              + polio.vacc 
              + HIV.AIDS 
              + Status, lifedata, nbest=3, nvmax=11)
              
plot(life.rs, scale='adjr2')
              
```


